Smart Model Elimination for Efficient Automated Machine Learning in Healthcare

### 2024-06-24 paper/paper.tex
variable importance
Update paper.tex

\begin{abstract}
  Automated Machine Learning or AutoML has emerged as a popular field of research. We performed a literature review of existing AutoML papers and conducted an in-depth survey on six popluar AutoML frameworks. Many of these frameworks optimize a form of model selection, however, they all require every model to be run and evaluated. We propose a novel framework, SMEML (Smart Model Elimination Machine Learning), that automatically eliminates models that are unlikely to be performant so unnecessary time is not spent training these models. SMEML demonstrates the ability to generate a model with near equal accuracy to a "dumb" brute-force model that runs every model on every dataset, but in a fraction of the time. We also believe this innovation is particularly to machine learning in healthcare because of the necessity for efficient diesease prediction. 
\end{abstract}

@@

One sector of the global industry that has the most potential for beneficial integration of AutoML frameworks is healthcare, particularly in developing countries. These countries have severe shortages of healthcare workers and limited tools for diagnosis. For example, Africa has 2.3 healthcare workers per 1000 individuals, while the Americas have 24.8 healthcare workers per 1000 [2]. The World Health Organization (WHO) emphasized that this deficit is growing every year and will likely reach 18 million personnel by 2030 [3]. The use of AI in healthcare has been varied. Medical AI's typically automate repetitive tasks, making time consumption a primary concern [4]. This coupled with a lack of adequate resources makes designing faster diagnostic systems for developing countries' medical sectors a pivotal issue. 
%
Most AutoML frameworks implement some form of model selection where a pool of models are filtered until a final model is selected. However, these frameworks require that every model to be run to filter out [1]. In theory, this means that much energy is spent training models that are unlikely to be selected. We propose SMEML, a novel model selection algorithm that automatically eliminates models that it believes will not be performant.

@@

\begin{table}
  \caption{Survey of Existing AutoML Frameworks}
  \label{automl-survey}
  \centering
  \begin{tabular}{lllllllll}
    \toprule
    Framework & EOU & M & FE & TM & HPT & Inter & Custom\\
    \midrule
    H2O AutoML & Easy & 6 & Basic & Yes & Auto & SHAP & High\\
    TPOT & Hard & 10 & Basic & No & Genetic & POJO & Moderate\\
    MLJAR & Easy & 12 & Advanced & Yes & Auto & Basic & High\\
    FLAML & Easy & 9 & Limited & Yes & Auto & SHAP & Low-Moderate\\
    LightAutoML & Easy & 11 & Advanced & Yes & Auto & Basic & Moderate\\
    \bottomrule
  \end{tabular}
  \captionsetup{font=footnotesize}
  \caption*{\textbf{Abbreviations}: EOU: Ease of Use, M: \# of models evaluated by framework, FE: Feature Engineering, TM: Time Management, HPT: Hyper-Parameter Tuning, Inter: Interface, Custom: Customization}
\end{table}
We systematically reviewed and compared five different AutoML frameworks, comparing every feature relevant to medical implementation [5-10]. Of these six, we wanted to take a closer look at four: FLAML, H2O, MLJAR, and TPOT.

Taking a closer look at four AutoML standards, we can observe pros and cons in their theoretical application to medicine and research. FLAML uses blend search hyper-parameter optimization and focuses on lightweight models, making it quick and applicable to repetitive tasks in the medical sphere. It supports imputation techniques for missing values with user specification and has an intuitive API, making it moderately user-friendly for minimally trained healthcare workers [9]. H2O has a wider variety of supported algorithms that may or may not be appropriate for medical diagnosis, making it slower than FLAML on average. In one trial, more than half of FLAML's performances in one minute were better than or equal to H20's performances in one hour [9]. H2O has automatic and multi-faceted methods of dealing with missing values and flexible interfaces in R and Python, making it intuitive for users [5]. MLJAR uses more advanced algorithms such as light gradient boosting and neural networks. It automatically deals with missing values and is known for its automated user interface [7]. TPOT uses genetic programming. It builds pipelines over multiple generations, which can be time consuming. It has built-in pre-processing for missing values, but the genetic programming can require fine-tuning from users [6].

Keeping the factors we have laid out in mind, namely expeditious running, accuracy, and interpretability, we also uncovered a key continuity in framework design: frameworks are forced to run and train the majority of their available models to maximize accuracy at the expense of efficiency. Given the restricted nature of tabular data, there are some models that, in general, are less likely to perform well, thus running and training these models is frequently wasteful. Cutting these models completely, however, is not optimal either, as they are the best option in a minority of cases and give frameworks versatility. We hypothesized that using a combination of layered machine learning implementation and meta-feature extraction would yield an equally versatile system that minimizes time spent on prediction while still maximizing the other quantities.

\section{Methodology}
When considering a specific dataset, various aspects regarding the distribution and relationship within the data can give us clues as to which models are likely to perform best. For example, linear models tend to work better on datasets where there is a linear relationship between features [11]. As a result, we can develop a boosting model to predict the best models based off extracted attributes from a dataset. These extracted attributes are shown in Table 2.

@@

After running the benchmark datasets, several observations were readily apparent. First, accuracy did not change for 4 of the 6 benchmarks. The other two benchmarks on eye states and diabetes prediction represented a loss of accuracy for SMEML by 0.13\% and 1.95\%, respectively. Training time, however, favored SMEML in every trial. SMEML outperformed the dummy framework by 173.96 seconds in its worst performance on the lung cancer trial and by 383.94 seconds in its best performance on the eye state dataset. Relative to the dummy's performance, a percent decrease in time to run for SMEML of at least 46.78\% could be seen, with a decrease of up to 69.22\% observerd as well. SMEML's worst performance by percent decrease was also the performance in which it returned perfect accuracy. 
\section{Discussion}
\subsection{Summary}
As evidenced by our experiment's results, SMEML manages to return comparable accuracy results compared to its dumb couterpart while more than splitting the run time in half in the majority of benchmarks. Its performance is not limited by the size of the dataset, and it in fact returned its first, third, and fourth best percentage-based run times on the 'large' dataset benchmarks (those exceeding 1000 rows). This is important in third world medicine where time is of the essence and there is a constant influx of new patient data. Further research should be done in incorporating this innovation into a full-fledged AutoML framework to expedite and improve the process further. A practical demonsration of this software's real-world application can be found on our SmartDx website [18]. Source code can be found at [19].
\subsection{Limitations}
Our implementation of SMEML has limitations. These include a small training set due to lack of available datasets, finite computing power, time constraints, and a lack of access to paticular scientific literature. Our framework only improves on the model selection process and does not address the various other aspects of AutoML. Therefore, SMEML is not comparable to other full-featured AutoML frameworks.
\subsection{Future Work}
Future work will include increasing the size of the training set, implementing a fully fledged AutoML framework, and conducting a detailed experiment comparing SMEML to other popular AutoML frameworks on more datasets. Furthermore, we can examine other types of machine learning such as regression and multiclass-classification.
\begin{ack}
The authors would like to thank the St. Mark's School of Texas for supporting this research.
\end{ack}

>>>>

\begin{abstract}
  Automated Machine Learning or AutoML has emerged as a popular field of research. We performed a literature review of existing AutoML papers and conducted an in-depth survey on five popluar AutoML frameworks. Many of these frameworks optimize a form of model selection, however, they all require every model to be run and evaluated. We propose a novel framework, SMEML (Smart Model Elimination Machine Learning), that automatically eliminates models that are unlikely to be performant. SMEML demonstrates the ability to generate a model with near equal accuracy to a "dumb" brute-force model, but in a fraction of the time. We also believe this innovation is particularly relevant to machine learning in healthcare because of the necessity for efficient diesease prediction. 
\end{abstract}

@@

One sector of the global industry that has the most potential for beneficial integration of AutoML frameworks is healthcare, particularly in developing countries. These countries have severe shortages of healthcare workers and limited tools for diagnosis. For example, Africa has 2.3 healthcare workers per 1000 individuals, while the Americas have 24.8 healthcare workers per 1000 [2]. The World Health Organization (WHO) emphasized that this deficit is growing every year and will likely reach 18 million personnel by 2030 [3]. Medical AI's typically automate repetitive tasks, making time consumption a primary concern [4]. This coupled with a lack of adequate resources makes designing faster diagnostic systems for developing countries' medical sectors a pivotal issue. 
%
Most AutoML frameworks implement some form of model selection where a pool of models are filtered until a final model is selected. However, these frameworks require that every model to be run to filter out [1]. In theory, this means that much energy is spent training models that are unlikely to be selected. We propose SMEML, a novel model selection algorithm that automatically eliminates models that it believes will not be performant.

@@

\begin{table}
  \caption{Survey of Existing AutoML Frameworks}
  \label{automl-survey}
  \centering
  \begin{tabular}{lllllllll}
    \toprule
    Framework & EOU & M & FE & TM & HPT & Inter & Custom\\
    \midrule
    H2O AutoML & Easy & 12 & Basic & Yes & Auto & SHAP & High\\
    TPOT & Hard & 10 & Basic & No & Genetic & POJO & Moderate\\
    MLJAR & Easy & 12 & Advanced & Yes & Auto & Basic & High\\
    FLAML & Easy & 9 & Limited & Yes & Auto & SHAP & Low-Moderate\\
    LightAutoML & Easy & 11 & Advanced & Yes & Auto & Basic & Moderate\\
    \bottomrule
  \end{tabular}
  \captionsetup{font=footnotesize}
  \caption*{\textbf{Abbreviations}: EOU: Ease of Use, M: \# of models evaluated by framework, FE: Feature Engineering, TM: Time Management, HPT: Hyper-Parameter Tuning, Inter: Interface, Custom: Customization}
\end{table}
We systematically reviewed and compared five different AutoML frameworks, comparing every feature relevant to medical implementation [5-10]. Of these six, we wanted to take a closer look at four: FLAML, H2O, MLJAR, and TPOT.

We can observe pros and cons in the four models' theoretical applications. FLAML uses blend search hyper-parameter optimization and focuses on lightweight models, making it quick and applicable to repetitive tasks. It supports user-specified imputation techniques for missing values and has an intuitive API, making it moderately user-friendly for healthcare workers [9]. H2O has a wider variety of supported algorithms that may be appropriate for medical diagnosis, making it slower than FLAML on average. In one trial, more than half of FLAML's performances in one minute were better than or equal to H20's performances in one hour [9]. H2O has automatic and multi-faceted methods of dealing with missing values and flexible interfaces in R and Python, making it intuitive for users [5]. MLJAR uses advanced algorithms such as light gradient boosting and neural networks. It automatically deals with missing values and is known for its automated user interface [7]. TPOT uses genetic programming. It builds pipelines over multiple generations, which can be time consuming. It has built-in pre-processing for missing values, but the genetic programming can require fine-tuning from users [6].

Keeping the factors we have laid out in mind, namely expeditious running, accuracy, and interpretability, we also uncovered a key continuity in framework design: frameworks are forced to run and train the majority of their available models to maximize accuracy at the expense of efficiency. Given the restricted nature of tabular data, there are some models that are less likely to perform well, thus running and training these models is frequently wasteful. Cutting these models completely, however, is not optimal either, as they are the best option in a minority of cases. We hypothesized that using a combination of layered machine learning implementation and meta-feature extraction would yield an equally versatile system that minimizes time spent on prediction while still maximizing the other quantities.

\section{Methodology}
When considering a specific dataset, various aspects regarding the distribution and relationship within the data can give us clues as to which models are likely to perform best. For example, linear models tend to work better on datasets where there is a linear relationship between features [11]. As a result, we can develop a boosting model to predict the best models based off extracted attributes from a dataset. These extracted attributes are shown in Table 2. Each attribute attempts to capture a specific type of relationship or distribution within the data. The variable importance of each attribute is shown in the "Weight" column. Linearity and data distribution are the most important attributes in determining the best model.

@@

Several observations were readily apparent from running these datasets. First, accuracy did not change for 4 of the 6 benchmarks. The other two benchmarks on eye states and diabetes prediction represented a loss of accuracy for SMEML by 0.13\% and 1.95\%, respectively. Training time, however, favored SMEML in every trial. SMEML outperformed the dummy framework by 173.96 seconds in its worst performance on the lung cancer trial and by 383.94 seconds in its best performance on the eye state dataset. Relative to the dummy's performance, a percent decrease in time to run for SMEML of at least 46.78\% could be seen, with a decrease of up to 69.22\% observerd as well. SMEML's worst performance by percent decrease was also the performance in which it returned perfect accuracy. 
\section{Discussion}
\subsection{Summary}
SMEML manages to return comparable accuracy results compared to its dumb couterpart while more than splitting the run time in half in the majority of benchmarks. Its performance is not limited by the size of the dataset, and it in fact returned its first, third, and fourth best percentage-based run times on the 'large' dataset benchmarks (those exceeding 1000 rows). This is important in third world medicine where time is of the essence and there is a constant influx of new patient data. Further research should be done in incorporating this innovation into a full-fledged AutoML framework to expedite and improve the process further. A practical demonsration of this software's real-world application can be found on our SmartDx website [18]. Source code can be found at [19].
\subsection{Limitations}
Our implementation of SMEML has limitations. These include a small training set due to lack of available datasets, finite computing power, time constraints, and a lack of access to paticular scientific literature. Our framework only improves on the model selection process and does not address the various other aspects of AutoML. Therefore, SMEML is not comparable to other full-featured AutoML frameworks.
\subsection{Future Work}
Future work will include increasing the size of the training set, implementing a fully fledged AutoML framework, and conducting a detailed experiment comparing SMEML to other popular AutoML frameworks. Furthermore, we can examine other types of machine learning such as regression and multiclass-classification.
\begin{ack}
The authors would like to thank the St. Mark's School of Texas for supporting this research.
\end{ack}
